### 2017-11-10 latex/mst_lukkarila_juri_2017.tex
edit

This work was carried out during the spring and summer of 2017 at the Speech Recognition research group of the Department of Signal Processing and Acoustics at Aalto University School of Electrical Engineering. This thesis would not have been possible without the support of the Academy of Finland for the project \textit{Conversation Assistant for the Hearing Impaired}. \\\\
First and foremost, I would like to thank supervisor Prof.\ Mikko Kurimo and my advisor D.Sc.\ Kalle Palomäki for giving me the opportunity to work on this interesting and multifaceted topic. I am grateful for the freedom and continued support given to me during this work. I would like to thank Symeon Delikaris-Manias and Juhani Paasonen from the Spatial Sound research group for assistance with the background noise recordings used in the user tests, and Ilkka Huhtakallio for providing the necessary audio equipment. My thanks to Olli Savisaari from the User Interfaces research group for consulting with user testing methods. My thanks to Seppo Enarvi, Reima Karhila, Katri Leino, Ulpu Remes, Aku Rouhe and Peter Smit for general help, tips and discussion. \\\\
I would like to thank Päivi Rainò from HUMAK and Tarja Kaikkonen from Kuuloliitto ry for helping to recruit deaf and hard of hearing persons for the user tests, and many thanks to the people who participated in the tests. \\\\
Finally, I would like to thank Siiri for all the support and motivation. \\

@@

Audiologically\footnote{Audiology is the scientific study of hearing, including the treatment of hearing defects.}, hearing loss is categorized and its severity ranked according to the increase in the threshold of hearing, i.e., the sound pressure level required for the perceptual detection of sound \cite{moore2007cochlear}. It is measured in decibels and compared to the statistically defined and standardized nominal level of hearing. Human hearing is strongly frequency-dependent, and is most sensitive at frequencies from approximately one to five kilohertz \cite{pulkki2015communication}. Consequently, this frequency range is critical for speech perception and many other everyday tasks. \\\\ Figure \ref{fig:loudness} presents the equal-loudness contours (curves) as they are defined in the ISO 226:2003 standard \cite{iso226}. These contours describe the \textit{sound pressure level} (SPL) required for a pure tone (single frequency sinusoidal waveform) to be judged equally loud depending on the frequency of the tone, illustrating the frequency-dependent sensitivity of hearing \cite{pulkki2015communication}. Notably, the equal-loudness curve for zero phons corresponds to the absolute threshold of hearing. Hearing loss can therefore be technically defined as upward changes to this curve, the frequency-dependent threshold of hearing, that exceed the normal small statistical variation between different persons. \\

@@

Hearing aids are used to amplify and modify incoming sound, enhancing and augmenting the users own compromised hearing \cite{levitt2007historical, salonen2013hearing, fink2008benefit}. Modern devices are based on digital electronics and microprocessors, and come in many different types. The most common variations are the different \textit{in-the-ear} and \textit{behind-the-ear} models. Figure \ref{fig:hearingaids} presents various types of modern hearings aids from one particular manufacturer (Widex). \\

@@

In addition to the simple amplification provided by earlier analog models, modern digital hearing aids utilize real-time \textit{digital signal processing} (DSP) for tasks such as speech enhancement, dynamics processing (compression), filtering, noise reduction, adaptive gain control and feedback cancellation \cite{moore2007cochlear, levitt2007historical, goehring2016speech}. Directional microphone systems are used for improving speech recognition in noise. Modern wireless transmission technology, such as Bluetooth, can be used to easily link hearing aids directly with sound sources like a telephone or television \cite{salonen2013hearing}. Modern hearing aids are personally fitted for each user, matching and compensating for the individual frequency response of the users hearing for best results \cite{moore2007cochlear}. \\\\

@@

	\caption{Block diagram for the structure of a typical speech recognition system.}

@@

	\includegraphics[trim={0cm 0.3cm 0.4cm 0.5cm}, clip, width=\textwidth]{audio.pdf}

@@

What is needed are some characteristic measures, or \textit{features}, that can describe and discriminate between different speech sounds, and extract the linguistic information independent of the speakers personal voice. These features should contain the essential information and measurements needed for classifying speech. \textit{Feature extraction} is then the process of calculating a sequence of these features, a \textit{feature vector}, based on the input audio signal. Ideally, a feature vector should be compact, containing only the necessary information, be robust against noise, as well as fast to calculate so that it can be used in real-time \cite{huang2001spoken}. Consequently, audio signals typically contain a lot of information that is not useful for this particular task. In feature extraction, the input signal is therefore processed to remove unwanted and redundant information as well as noise, while emphasizing the important characteristics. \\

@@

\end{equation} where $f$ is the frequency in Hertz. The anchor point is set to 1000 Hz $\leftrightarrow$ 1000 mel \cite[p.~174--175]{pulkki2015communication}. As we are interested in the frequency contents of the signal, each frame is windowed and converted to the frequency domain by calculating the short-time power spectrum \cite{gales2008application}. A mel-scale filter bank with triangular band-pass filters spaced in perceptually equally long frequency bands is applied to each frame. A mel-scale filter bank is presented in figure \ref{fig:melbank}. The combined energy of each frequency band is then calculated. Finally, the cepstral coefficients are obtained by first taking the logarithm of each energy band, mimicking the logarithmic loudness perception of hearing, and applying the discrete cosine transform, in effect compressing and decorrelating the energies \cite{huang2001spoken, gales2008application}. \\\\
The cepstrum can be viewed as the spectrum of a spectrum, as the inverse (discrete) Fourier transform of a log spectrum transforms to the cepstral domain.

@@

	\includegraphics[trim={0cm 0cm 0cm 0.5cm}, clip, width=\textwidth]{melfilter.pdf}

@@

	\includegraphics[trim={0.1cm 0cm 0cm 0cm}, clip, width=\textwidth]{cepstrum.pdf}

@@

\textit{Hidden Markov Models} (HMM) have been widely used for implementing the acoustic model \cite{gales2008application, hori2013speech}. Speech is approximated well by a \textit{Markov chain}, a stochastic model for randomly changing systems where the next state depends only on the current state \cite[p.~23--26]{yu2014automatic}. The chain consists of discrete states and state changes that are modeled with transition probabilities. In a hidden Markov model, the states cannot be observed directly (hence the term \textit{hidden}). Instead, an output that is dependent on the states is known. In speech recognition, an utterance of a phoneme is considered a hidden state that emits an observable representation in the form of a feature vector. A three-state HMM is generally used to model one triphone in a left-to-right topology, meaning the states are constrained to be in chronological order \cite{kallasjoki2016}. Figure \ref{fig:hmm} present a three-state, left-to-right HMM.

@@

	\includegraphics[width=\textwidth]{hmm.pdf}

@@

The probability that an observation is an emission of a specific state needs to be also modeled. For representing the probability distribution of observations, \textit{Gaussian Mixture Models} (GMM) have been a popular choice \cite{gales2008application, hinton2012deep}. Gaussian mixture models are built by linearly combining two or more multivariate Gaussian distributions together. A simplified example of an GMM is presented in figure \ref{fig:gmm}.

@@

	\includegraphics[trim={0cm 0.2cm 0cm 0.2cm}, clip, width=\textwidth]{gmm.pdf}

@@

The basic operation principle of the proposed Conversation Assistant can be described as follows: First, the speech of the speaker(s) is picked up with a microphone. This can be for example the build-in microphone of a users device, or an external microphone, preferably close to the speaker for improved signal-to-noise ratio and ASR performance. The speech input is then converted into text by an automatic speech recognition system. Finally, the recognition result is displayed on a screen for the user or users. All of these basic components can be separate from each other, or integrated into one device. The resulting text can be displayed with special visual formatting depending on additional features the system has, such as speaker recognition, enabling visual separation of speech from different speakers. \\\\
Speech recognition can be a computationally intensive task, and large models can require a lot of disk space as well as memory during operation. As such, most laptop computers can run a speech recognizer locally, but current mobile devices and less powerful computers cannot. Also, it is not practical or sensible that every user would need to download potentially multiple gigabytes of data to their own device in order for the speech recognition to happen on their device locally, even if the computational resources were sufficient. Instead, a server-based solution is needed, where the audio signal is acquired at the end-device and is then send via the internet to a server handling the speech recognition process. The server then simply returns the recognition results as a text string to the device for display. This is the way most ASR systems aimed for general consumers currently work already. The major downside of the server-based implementation is the required internet connection. \\\\
Mobile devices \cite{mcgraw2016personalized}.

@@

\subsection{Numerical values}

The numerical data obtained, i.e., the numerical ratings given by the test participants, are presented with a box plot for each question, divided by the section. The actual numbers given by the test participants for each question are included in appendix \ref{sec:answers}, and the MATLAB code in appendix \ref{sec:matlab}. In the box plots, the (red) central mark represents the median value. The bottom of the box (in blue) corresponds to the 25th percentile, and the top of the box to the 75th percentile. It should be noted that when the median is not centered in the box, it shows sample skewness, meaning the values are distributed asymmetrically. The lines (in black) extending from the box display the range of all values that are not considered outliers. A value is judged to be an outlier and marked with a red cross if it is more than 1,5 times the interquartile range (i.e., the size of the box) away from the top or bottom of the box. \cite{boxplot} \\\\

@@

More testing. Better models for conversation speech. As they say: "There's no data like more data." \\\\
Group testing. \\\\
How useful depending on level of hearing impairment! \\\\
Web and mobile application, server-based recognizer. \\\\
Microphone methods: access other peoples mobile devices in meetings. Beamforming and directional long range mics. \\\\
Speaker diarization. \\\\
Augmented reality.

@@

\section{MATLAB Code} \label{sec:matlab}

This section presents the MATLAB scripts created for analyzing the user test results and producing most of the figures presented in this thesis.

\lstset{
	language=matlab,
	basicstyle=\ttfamily\scriptsize, % \tiny 
	inputencoding=utf8,
	extendedchars=true,
}

\subsection*{results.m}
\vspace{-4mm}
\lstinputlisting{results.m}
\vspace{8mm}
\subsection*{audiosignal.m}
\vspace{-4mm}
\lstinputlisting{audiosignal.m}
\vspace{8mm}
\subsection*{bformat.m}
\vspace{-4mm}
\lstinputlisting{bformat.m}
\vspace{8mm}
\subsection*{cepstrum.m}
\vspace{-4mm}
\lstinputlisting{cepstrum.m}
\vspace{8mm}
\subsection*{gmm.m}
\vspace{-4mm}
\lstinputlisting{gmm.m}
\vspace{8mm}
\subsection*{filterbank.m}
\vspace{-4mm}
\lstinputlisting{filterbank.m}
\vspace{8mm}
\subsection*{loudness.m}
\vspace{-4mm}
\lstinputlisting{loudness.m}

>>>>

This work was carried out during the spring and summer of 2017 at the Speech Recognition research group of the Department of Signal Processing and Acoustics at Aalto University School of Electrical Engineering. This thesis would not have been possible without the support of the Academy of Finland for the project \textit{Conversation Assistant for the Hearing Impaired}. \\\\
First and foremost, I would like to thank supervisor Prof.\ Mikko Kurimo and my advisor D.Sc.\ Kalle Palomäki for giving me the opportunity to work on this interesting and multifaceted topic. I am grateful for the freedom and continued support given to me during this work. I would like to thank Symeon Delikaris-Manias and Juhani Paasonen from the Spatial Sound research group for assistance with the background noise recordings used in the user tests, and Ilkka Huhtakallio for providing the necessary audio equipment. My thanks to Olli Savisaari from the User Interfaces research group for consulting with user testing methods. From the Speech Recognition group, I would like to thank Seppo Enarvi, Reima Karhila, Katri Leino, Ulpu Remes, Aku Rouhe and Peter Smit for general help, tips and discussion. \\\\
I would like to thank Päivi Rainò from HUMAK and Tarja Kaikkonen from Kuuloliitto ry for helping to recruit deaf and hard of hearing persons for the user tests, and many thanks to the people who participated in the tests. \\\\
Finally, I would like to thank Siiri for all the support and motivation. \\

@@

Audiologically\footnote{Audiology is the scientific study of hearing, including the treatment of hearing defects.}, hearing loss is categorized and its severity ranked according to the increase in the threshold of hearing, i.e., the sound pressure level required for the perceptual detection of sound \cite{moore2007cochlear}. It is measured in decibels and compared to the statistically defined and standardized nominal level of hearing. Human hearing is strongly frequency-dependent, and is most sensitive at frequencies from approximately one to five kilohertz \cite{pulkki2015communication}. Consequently, this frequency range is critical for speech perception and many other everyday tasks. Figure \ref{fig:loudness} presents the equal-loudness contours (curves) as they are defined in the ISO 226:2003 standard \cite{iso226}. These contours describe the \textit{sound pressure level} (SPL) required for a pure tone (single frequency sinusoidal waveform) to be judged equally loud depending on the frequency of the tone, illustrating the frequency-dependent sensitivity of hearing \cite{pulkki2015communication}. Notably, the equal-loudness curve for zero phons corresponds to the absolute threshold of hearing. Hearing loss can therefore be technically defined as upward changes to this curve, the frequency-dependent threshold of hearing, that exceed the normal small statistical variation between different persons. \\

@@

Hearing aids are used to amplify and modify incoming sound, enhancing and augmenting the users own compromised hearing \cite{levitt2007historical, salonen2013hearing, fink2008benefit}. Modern devices are based on digital electronics and microprocessors, and come in many different types. The most common variations are the different \textit{in-the-ear} and \textit{behind-the-ear} models. Figure \ref{fig:hearingaids} presents various types of modern hearings aids from one particular manufacturer (Widex). In addition to the simple amplification provided by earlier analog models, modern digital hearing aids utilize real-time \textit{digital signal processing} (DSP) for tasks such as speech enhancement, dynamics processing (compression), filtering, noise reduction, adaptive gain control and feedback cancellation \cite{moore2007cochlear, levitt2007historical, goehring2016speech}. Directional microphone systems are used for improving speech recognition in noise. Modern wireless transmission technology, such as Bluetooth, can be used to easily link hearing aids directly with sound sources like a telephone or television \cite{salonen2013hearing}. Modern hearing aids are personally fitted for each user, matching and compensating for the individual frequency response of the users hearing for best results \cite{moore2007cochlear}. \\

@@

@@

	\caption{Block diagram for the structure of a typical speech recognition system. The audio input can be a real-time signal from a microphone, or a existing recording or audio track. Depending on the application, the text result can be displayed immediately or saved into a text file.}

@@

	\begin{subfigure}[b]{\textwidth}
		\includegraphics[trim={0.4cm 0cm 0.1cm 0.2cm}, clip, width=\textwidth]{waveform.pdf}
	\end{subfigure}
	\begin{subfigure}[b]{\textwidth}
		\includegraphics[trim={0.4cm 0cm 0.1cm 0cm}, clip, width=\textwidth]{spectrogram.pdf}
	\end{subfigure}

@@

What is needed are some characteristic measures, or \textit{features}, that can describe and discriminate between different speech sounds, and extract the linguistic information independent of the speakers personal voice. These features should contain the essential information and measurements needed for classifying speech. \textit{Feature extraction} is then the process of calculating a sequence of these features, a \textit{feature vector}, based on the input audio signal. Ideally, a feature vector should be compact, containing only the necessary information, be robust against noise, as well as fast to calculate so that it can be used in real-time \cite{huang2001spoken}. Consequently, audio signals typically contain a lot of information that is not useful for this particular task. In feature extraction, the input signal is therefore processed to remove unwanted and redundant information as well as noise, while emphasizing the important characteristics. \\\\

@@

\end{equation} where $f$ is the frequency in Hertz. The anchor point is set to 1000 Hz $\leftrightarrow$ 1000 mel \cite[p.~174--175]{pulkki2015communication}. As we are interested in the frequency contents of the signal, each frame is windowed and converted to the frequency domain by calculating the short-time power spectrum \cite{gales2008application}. A mel-scale filter bank with triangular band-pass filters spaced in perceptually equally long frequency bands is applied to each frame. A mel-scale filter bank is presented in figure \ref{fig:melbank}. The combined energy of each frequency band is then calculated. Finally, the cepstral coefficients are obtained by first taking the logarithm of each energy band, mimicking the logarithmic loudness perception of hearing, and applying the discrete cosine transform, in effect compressing and decorrelating the energies \cite{huang2001spoken, gales2008application}. The cepstrum can be viewed as the spectrum of a spectrum, as the inverse (discrete) Fourier transform of a log spectrum transforms to the cepstral domain.

@@

	\includegraphics[trim={0.1cm 0cm 0cm 0.5cm}, clip, width=\textwidth]{melfilter.pdf}

@@

	\includegraphics[trim={0.1cm 0cm 0cm 0.4cm}, clip, width=\textwidth]{cepstrum.pdf}

@@

\textit{Hidden Markov Models} (HMM) have been widely used for implementing the acoustic model \cite{gales2008application, hori2013speech}. Speech is approximated well by a \textit{Markov chain}, a stochastic model for randomly changing systems where the next state depends only on the current state \cite[p.~23--26]{yu2014automatic}. 

@@

	\includegraphics[trim={0.1cm 0cm 0.1cm 0.5cm}, clip, width=\textwidth]{hmm2.pdf}

@@

The chain consists of discrete states and state changes that are modeled with transition probabilities. In a hidden Markov model, the states cannot be observed directly (hence the term \textit{hidden}). Instead, an output that is dependent on the states is known. In speech recognition, an utterance of a phoneme is considered a hidden state that emits an observable representation in the form of a feature vector. A three-state HMM is generally used to model one triphone in a left-to-right topology, meaning the states are constrained to be in chronological order \cite{kallasjoki2016}. Figure \ref{fig:hmm} present a three-state, left-to-right HMM. The probability that an observation is an emission of a specific state needs to be also modeled. For representing the probability distribution of observations, \textit{Gaussian Mixture Models} (GMM) have been a popular choice \cite{gales2008application, hinton2012deep}. Gaussian mixture models are built by linearly combining two or more multivariate Gaussian distributions together. A simplified example of an GMM is presented in figure \ref{fig:gmm}.

@@

	\includegraphics[trim={0cm 0.2cm 0cm 0.2cm}, clip, width=\textwidth]{gmm2.pdf}

@@

The basic operation principle of the proposed Conversation Assistant can be described as follows: First, the speech of the speaker(s) is picked up with a microphone. This can be for example the build-in microphone of a users device, or an external microphone, preferably close to the speaker for improved signal-to-noise ratio and ASR performance. The speech input is then converted into text by an automatic speech recognition system. Finally, the recognition result is displayed on a screen for the user or users. All of these basic components can be separate from each other, or integrated into one device. The resulting text can be displayed with special visual formatting depending on additional features the system has, such as speaker recognition, enabling visual separation of speech from different speakers. \\\\
Speech recognition can be a computationally intensive task, and large models can require a lot of disk space as well as memory during operation. As such, most laptop computers can run a speech recognizer locally, but current mobile devices and less powerful computers cannot. Also, it is not practical or sensible that every user would need to download potentially multiple gigabytes of data to their own device in order for the speech recognition to happen on their device locally, even if the computational resources were sufficient. Instead, a server-based solution is needed, where the audio signal is acquired at the end-device and is then send via the internet to a server handling the speech recognition process. The server then simply returns the recognition results as a text string to the device for display. This is the way most ASR systems aimed for general consumers currently work already. The major downside of the server-based implementation is the required internet connection. \\\\
Mobile devices \cite{mcgraw2016personalized}. \\\\
\begin{figure}[h]
	\centering
	\includegraphics[trim={0cm 0cm 0cm 0cm}, clip, width=\textwidth]{implementation.pdf}
	\caption{The Conversation Assistant system.}
	\label{fig:system} 
\end{figure}

@@

\subsection{Numerical ratings}

The numerical data obtained, i.e., the numerical ratings given by the test participants, are presented with a box plot for each question, divided by the section. The actual numbers given by the test participants for each question are included in appendix \ref{sec:answers}. In the box plots, the (red) central mark represents the median value. The bottom of the box (in blue) corresponds to the 25th percentile, and the top of the box to the 75th percentile. It should be noted that when the median is not centered in the box, it shows sample skewness, meaning the values are distributed asymmetrically. The lines (in black) extending from the box display the range of all values that are not considered outliers. A value is judged to be an outlier and marked with a red cross if it is more than 1,5 times the interquartile range (i.e., the size of the box) away from the top or bottom of the box. \cite{boxplot} \\\\

@@

More testing. Better models for conversation speech. As they say: "There's no data like more data." \\\\
Group testing. \\\\
How useful depending on level of hearing impairment! \\\\dr
Web and mobile application, server-based recognizer. \\\\
Microphone methods: access other peoples mobile devices in meetings. Beamforming and directional long range mics. \\\\
Speaker diarization. \\\\
Augmented reality.

@@

%\section{MATLAB Code} \label{sec:matlab}

%This section presents the MATLAB scripts created for analyzing the user test results and producing most of the figures presented in this thesis.

%\lstset{
%	language=matlab,
%	basicstyle=\ttfamily\scriptsize, % \tiny 
%	inputencoding=utf8,
%	extendedchars=true,
%}

%\subsection*{results.m}
%\vspace{-4mm}
%\lstinputlisting{results.m}
%\vspace{8mm}
%\subsection*{audiosignal.m}
%\vspace{-4mm}
%\lstinputlisting{audiosignal.m}
%\vspace{8mm}
%\subsection*{bformat.m}
%\vspace{-4mm}
%\lstinputlisting{bformat.m}
%\vspace{8mm}
%\subsection*{cepstrum.m}
%\vspace{-4mm}
%\lstinputlisting{cepstrum.m}
%\vspace{8mm}
%\subsection*{gmm.m}
%\vspace{-4mm}
%\lstinputlisting{gmm.m}
%\vspace{8mm}
%\subsection*{filterbank.m}
%\vspace{-4mm}
%\lstinputlisting{filterbank.m}
%\vspace{8mm}
%\subsection*{loudness.m}
%\vspace{-4mm}
%\lstinputlisting{loudness.m}
